---
title: "Tutorial for identifying genes with spatial expression trends using trendsceek"
author: "Daniel Edsgärd"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial for identifying genes with spatial expression trends using trendsceek}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Synthetic data

```{r }
##Load library
library('trendsceek')
```

### Create syntethic data
Below we simulate a step, hotspot and streak of cells with higher expression than the background.

```{r }
##Set parameters
##approximate number of cells
nsim_cells = 300

##low and high expression levels
low_expr = c(10, 10)
high_expr = c(20, 50)

##Randomly position the cells in 2d space
pp = sim_pois(nsim_cells)

##Add gene expression levels as mark distributions
pp = add_markdist_step(pp, low_expr, high_expr)
pp = add_markdist_hotspot(pp, low_expr, high_expr)
pp = add_markdist_streak(pp, low_expr, high_expr)
```

Plot synthetic data

```{r sim-scatter-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp, log_marks = FALSE, scale_marks = FALSE, pal.direction = -1)

```{r sim-density-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_density(pp, log_marks = FALSE)
```

### Run trendsceek

```{r sim-trendsceek, results='hide'}
##set parameters
nrand = 100
ncores = 1

##run
trendstat_list = trendsceek_test(pp, nrand, ncores)

##extract significant genes
alpha = 0.1 ##Benjamini-Hochberg
sig_list = extract_sig_genes(trendstat_list, alpha)
```

Plot trend-statistics for significant genes in synthetic dataset

```{r sim-trendstats-plot, fig.width = 7, fig.height = 7, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
sig_genes = sig_list[['markcorr']][, 'gene']
plot_trendstats(trendstat_list, sig_genes)
```

The grey band indicates a null-distribution with Bonferroni-adjusted limits and may therefore be more conservative than the Benjamini-Hochberg P-value used to extract the significant genes.

```{r real-data}
```

## Real data
### Data preparation
#### Read data

```{r }
data('scialdone')
counts = scialdone[['counts']]
```

#### Gene variability stats
Filter genes on being expressed

```{r }
##Pass genes having at least 3 cells with minimum 5 in read count
min.ncells.expr = 3
min.expr = 5
counts_filt = genefilter_exprmat(counts, min.expr, min.ncells.expr)
dim(counts_filt)
```

Calculate gene variability stats

```{r }
quantile.cutoff = 0.9 ##filter out the most lowly expressed genes from the fitting
method = 'glm' ##For (robust) linear regression set to 'rlm'
vargenes_stats = calc_varstats(counts_filt, counts_filt, quant.cutoff = quantile.cutoff, method = method)
```

Select subset with the top-ranked most variable genes

```{r }
n.topvar = 500
topvar.genes = rownames(vargenes_stats[['real.stats']])[1:n.topvar]
```

Plot gene variability stats, highlighting the 500 most variable genes

```{r real-varstats-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot.ercc.points = FALSE
plot_cv2vsmean(vargenes_stats, topvar.genes, plot.ercc.points = plot.ercc.points)
```

#### Normalize the read counts

```{r }
min.count = 1
counts_norm = deseq_norm(counts, min.count)
```

#### Subset normalized counts on the most variable genes

```{r }
counts_sub = counts_norm[topvar.genes, ]
dim(counts_sub)
```

#### tSNE dimensionality reduction

```{r }
##params
tsne.k = 2 #Number of tSNE dimensions
init.dims = 100 #Number of PCA-dimensions to reduce to before tSNE
perp.frac = 0.2 #Fraction of cells 
max.iter = 300 #Maximum number of iterations
epoch = 50 #print status every epoch iteration

##run tsne including initial pca
tsne_res = trend.tsne(counts_sub, tsne.k, init.dims, perp.frac, max.iter, epoch)
```

If the error does not seem to have converged increase the argument 'max.iter'

```{r create-point-pattern}
```

#### Create point-pattern using tSNE positions as spatial distribution and expression levels as mark distribution

```{r }
##Convert tSNE cell positions to point pattern
pp = pos2pp(tsne_res)

##Set marks as the logged normalized gene expression 
log.fcn = log10
pp = set_marks(pp, counts_sub, log.fcn = log.fcn)
```

#### Plot top variable genes in tSNE space

```{r }
##Subset on top variable genes
n.top2plot = 10
topvar.genes = rownames(vargenes_stats[['real.stats']])[1:n.top2plot]
pp2plot = pp_select(pp, topvar.genes)

```{r real-tsne-scatter-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp2plot, log_marks = FALSE, scale_marks = FALSE, pal.direction = -1)
```

A number of the most variable genes do not seem to have a spatial trend, e.g. Dqx1 and the Gm* genes.
To find genes that have a spatial trend in an automated fashion and to assess the significance of the spatial trend we run trendsceek.

```{r real-run-trendsceek}
```

### Run trendsceek
To decrease the time it takes to run this vignette-example we run trendsceek using just the 10 most variable genes and with 100 permutations.

```{r real-trendsceek, results='hide'}
##set parameters
nrand = 100
ncores = 1

##run
trendstat_list = trendsceek_test(pp2plot, nrand, ncores)
```

### Analyze trendsceek results
A table with gene test-statistics and p-values is output by trendsceek_test.

```{r }
head(trendstat_list[['supstats_wide']])
```

Extract significant genes

```{r }
alpha = 0.05 ##Benjamini-Hochberg
sig_list = extract_sig_genes(trendstat_list, alpha)
```

Number of significant genes

```{r }
lapply(sig_list, nrow)
```

#### Significant mark-correlation genes

```{r }
##get significant genes
sig_genes = sig_list[['markcorr']][, 'gene']
```

Plot trend-statistics for significant genes

```{r real-markcorr-trendstats-plot, fig.width = 7, fig.height = 7, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_trendstats(trendstat_list, sig_genes)
```

Create marked point-pattern containing subset of significant genes

```{r }
pp_sig = pp_select(pp, sig_genes)
```

Scatter-plot with unscaled expression levels

```{r real-markcorr-scatter-notscaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp_sig, log_marks = FALSE, scale_marks = FALSE, pal.direction = -1)
```

Plots with scaled expression per gene

```{r real-markcorr-scatter-scaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp_sig, log_marks = FALSE, scale_marks = TRUE, pal.direction = -1)

```{r real-markcorr-density-scaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_density(pp_sig, log_marks = FALSE)
```

#### Identify and highlight cells located in regions exceeding background level
The function "cellsceek_test" creates a (cell x gene) table with values of 1 if a cell is in a region exceeding the background level and 0 otherwise.

```{r real-celltest, results='hide', message=FALSE, warning=FALSE}
nrand = 100
cellpeaks_siggenes = cellsceek_test(pp_sig, nrand = nrand)
sig_cells = get_sigcells(cellpeaks_siggenes)

```{r real-celltest-sigcells}
head(sig_cells, 10)
```

Plot position of cells located in regions exceeding the null distribution background level

```{r real-sigcells-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_density(pp_sig, log_marks = FALSE, cells2highlight = sig_cells)
```

#### Significant mark-variogram genes

```{r }
##get significant genes
sig_genes = sig_list[['markvario']][, 'gene']
```

Plot trend-statistics for significant genes

```{r real-markvario-trendstats-plot, fig.width = 7, fig.height = 7, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_trendstats(trendstat_list, sig_genes)
```

Create marked point-pattern containing subset of significant genes

```{r }
pp_sig = pp_select(pp, sig_genes)
```

Scatter-plot with unscaled expression levels

```{r real-markvario-scatter-notscaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp_sig, log_marks = FALSE, scale_marks = FALSE, pal.direction = -1)
```

Plots with scaled expression per gene

```{r real-markvario-scatter-scaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_scatter(pp_sig, log_marks = FALSE, scale_marks = TRUE, pal.direction = -1)

```{r real-markvario-density-scaled-plot, fig.width = 5, fig.height = 5, include = TRUE, echo = TRUE, fig.align = 'left', fig.cap = ''}
plot_pp_density(pp_sig, log_marks = FALSE)
```

